---
title: "Pelican classifcation"
format: html
editor: source
---

## 

https://towardsdatascience.com/data-science-tutorials-training-a-random-forest-in-r-a883cc1bacd1

# Random forests with ranger

```{r}
#install.packages("ranger")

library(tidyverse)
library(ranger)

source("R/confusion.R")

```

## RFs with ranger

Training testing example

```{r}
ranger(Species ~ ., data = iris)

train.idx <- sample(nrow(iris), 2/3 * nrow(iris))
iris.train <- iris[train.idx, ]
iris.test <- iris[-train.idx, ]

fit_iris <- ranger(Species ~ ., data = iris.train)

evaluate_model(fit_iris , iris.train, iris.test, y = "Species")
```

## With pelicans

Load data

```{r}

peli_labelled_raw <- read_csv("data/trainingData.csv")

peli_labelled <- 
  peli_labelled_raw |> 
  mutate(
    is_pelican = ifelse(Class==2, "pelican", "background") |> as.factor()) |>
  select(is_pelican, everything()) |> 
  select(-Class)

```

Make training and testing

```{r}

npoints <- nrow(peli_labelled)
train_ids <- sample(npoints, 0.75 * npoints)

peli_labelled_train <- peli_labelled |> slice(train_ids)
peli_labelled_test <-  peli_labelled |> slice(-train_ids)
```

Single covariates

```{r}
fit_rf1 <- ranger(is_pelican ~ b1, data = peli_labelled_train)

evaluate_model(fit_rf1, peli_labelled_train, peli_labelled_test)
```

3 covariate - RGB

```{r}
fit_rf3 <- ranger(is_pelican ~ b1 + b2 + b3, data = peli_labelled_train)

evaluate_model(fit_rf3, peli_labelled_train, peli_labelled_test)
```

RF all covars

```{r}
fit_rf1 <- ranger(is_pelican ~ ., data = peli_labelled_train)

evaluate_model(fit_rf1, peli_labelled_train, peli_labelled_test)

```

Variable importance

```{r}
fit_rf <- ranger(is_pelican ~ ., data = peli_labelled_train, importance = "permutation")

importance(fit_rf)
```

Reduced model (top 5 covariates)

```{r}

top_vars <- importance(fit_rf) |> sort(decreasing = T) |> head(5) |> names() 

fit_rf_top <- 
  ranger(is_pelican ~ ., 
         data = peli_labelled_train |> select(is_pelican, any_of(top_vars))
         )

evaluate_model(fit_rf_top, peli_labelled_train, peli_labelled_test)
```

# Extension

Is many trees (the random forest) better than single decision tree?

```{r}
fit_rf_single <- ranger(is_pelican ~ ., data = peli_labelled_train,   num.trees =1)

evaluate_model(fit_rf_single, peli_labelled_train, peli_labelled_test)

```

Is RF better than regression (regression only works with binary outcome)

```{r}

fit_glm <- glm(is_pelican ~ ., data = peli_labelled_train, family = binomial)

evaluate_model(fit_glm, peli_labelled_train, peli_labelled_test)

```

# Now you can use the model on a big dataset

```{r}

# fit final model, with 100 trees rather default of 500 (faster with fewer trees, and 100 seems like enough here)
fit_rf1 <- ranger(is_pelican ~ ., data = peli_labelled_train, num.trees = 100)

# load mosaic to predict onto
data_mosaic <- readRDS("data/large-files/RasterWithPredictorsPelicans-0000011776-0000011776.rds") 

# make predictions
data_pred <- 
  data_mosaic |>
  mutate(is_pelican = predict(fit_rf1, data = data_mosaic)$predictions)

# Plot direct to file "output.png", works better than plotting to screen for big datastes 
png("output.png", height= 40, width=40, units = "cm", res=600)
ggplot(data_pred, aes(x,y)) +
  geom_tile(aes(fill=is_pelican))
dev.off()

```