---
title: "Week 1 First Practical"
author: 
format: html 
editor: visual
---

# Week 1 First Practical

## Set up R packages

```{r}
# install.packages(c("readr", "ggplot2"))
library(readr)
library(ggplot2)
```

## Working with your own data

> Note: The quiz questions on Moodle are the same as the ones in this Quarto file. Please run the code in this file on Rstudio and answer the questions in Moodle as you go.

## Stalagmite data

Stalagmites are mineral formations that grow in caves, preserving environmental records in their layered structure. Paleoclimatologists from the [Climate Change Research Centre](https://www.unsw.edu.au/research/ccrc) collected this data. The thickness of each lamina reflects past climate conditions—thicker layers often indicate wetter periods, while thinner ones suggest drier times. By analysing lamina thickness and age, scientists can reconstruct past climate patterns, track seasonal variations, and identify historical environmental events.\
\
This dataset contains laminated stalagmite measurements, with 40,279 records. It includes the following columns:

-   `entity_id`: Identifier for the stalagmite sample.
-   `seq_id`: Sequence identifier (likely referring to sample grouping).
-   `lamina thickness (mm)`: Thickness of individual laminae in millimeters.
-   `lam_age (BP, 0=1950AD)`: Age of the lamina in years before present (BP), with 0 corresponding to 1950 AD.

```{r}
stalagmite_data <- read_csv("data/Laminated stalagmite Dataset.csv")
```

```{r}
stalagmite_data |> janitor::clean_names()
```

> You may find it helpful to use the function `clean_names()` from the `janitor` R package first before delving in. You may need to install this package first. Remember to use `library(janitor)` to load this package into R, put this code in the first R chunk!

Using the functions you've learned so far in this prac, try answer the following questions: 

**Questions:**

**How many rows and columns are in `stalagmite_data`?**

```{r}

```

**What type of variable is lamina thickness?**

```{r}

```

**What type of variable is `lam_age (BP, 0=1950AD)`?**

```{r}

```

**What is a more suitable variable type for `lam_age (BP, 0=1950AD)`?**

```{r}

```

**Write some code to change the variable type of `lam_age (BP, 0=1950AD)`?**

```{r}

```

**Write some code to display a quick summary of this dataset**

```{r}

```

## Sydney Beaches

Next, we'll do a data visualisation to better understand some patterns in our data. For this section, we'll be using the data frame `sydneybeaches.csv`

We want to try answer the question **which beach is the cleanest in Sydney?** (hint: the site with the least bacteria)

First, read in the data.

```{r}

```

**Check the structure to make sure the variable types were assigned correctly.**

```{r}

```

<!-- TODO multichoice -->

**Which of these variables were assigned correctly? (choose all that apply)**

-   `Region`

-   `Council`

-   `Latitude`

-   `Site`

We are primarily interested in the variables `Site` and `Enterococci_cfu_per_100ml`

It looks like `Enterococci_cfu_per_100ml` was assigned incorrectly.

**Change `Enterococci_cfu_per_100ml` to an integer variable.**

```{r}

```

Use this newly formatted data to visualise the amount of bacteria found at each `Site.`

<!-- TODO: R cloze question -->

To make the plot, replace the terms `your_data`, `x_axis`, and `y_axis` with something relevant.

```{r}
ggplot(your_data, aes(x_axis, y_axis)) +
         geom_boxplot() +
         theme(axis.text.x = element_text(angle = 90, size = 7))
```

**Questions:**

**Which Sydney beach is the cleanest?**

