---
title: "Week 1 First Practical"
author: 
format: html 
editor: visual
---

# Week 1 First Practical

## Set up R packages

```{r}
# install.packages(c("readr", "ggplot2"))
library(readr)
library(ggplot2)
```

## Working with your own data

> Note: The quiz questions on Moodle are the same as the ones in this quarto file. Please run the code in this file on Rstudio and answer the questions in Moodle as you go.

Until now, we've been working with data that is built into R. Now, we are going to try to explore our own data. Since R doesn't automatically have access to data that you have created, we will have to read it into our work space as an object. We can do this with `read_csv()`.

## Stalagmite data

Up until now

```{r}
dat <- read_csv("data/Laminated stalagmite Dataset.csv") # read in data 
```

(note: We used the same name for my data as the iris dataset. so this new data will replace the iris with the staalgmite data. It's ok to do this, so long as you mean to. People often name data table with a similar name like `dat` or `data`)

Look at the structure of the data set:

**How many rows and columns are there?**

```{r}

```

\[ answer \]

**What type of variable is lamina thickness?**

\[ answer \]

**Is this the correct variable type?**

\[ answer \]

**What type of variable is lam_age?**

\[ answer \]

This is not the correct variable type for this analysis.

**What should the correct variable type be?**

\[ answer \]

**How would you change it to the correct variable type?**

```{r}

```

`entity_id` and `seq_id` are both identifiers that have a limited number of unique strings. This means they should be coded as factors in the data set, but since they are numbers, they have been assigned the `num` variable type.

Change both variables to factors.

```{r}

```

Double check that your variables are all correct now using the structure function.

```{r}

```

## Sydney Beaches

Next, we'll do a quick visualization to better understand some patterns in our data.

For this section, we'll be using the data frame `sydneybeaches.csv`"that is in your working directory to answer our main question, which beach is the cleanest in Sydney? (hint: the site with the least bacteria)

First, read in the data.

```{r}

```

Check the structure to make sure the variable types were assigned correctly.

```{r}

```

Which of these variables were assigned correctly? (choose all that apply)

-   Region

-   Council

-   Latitude

-   Site

For our question we are interested in the variables `Site` and `Enterococci_cfu_per_100ml`

It looks like both of these have been assigned incorrectly.

Change `Site` to a factor variable and `Enterococci_cfu_per_100ml` to an integer variable.

```{r}

```

Use this format to visualize the amount of Enterococci bacteria found at each Site.\

To make the plot, replace the terms `your_data`, `x_axis`, and `y_axis` with something relevant.

```{r}
ggplot(your_data, aes(x_axis, y_axis)) +
         geom_boxplot() +
         theme(axis.text.x = element_text(angle = 90, size = 7))

```

**Which beach seems to have the least bacteria?**

-   South Maroubra Beach

-   Malabar Beach

-   Coogee Beach

-   Tamarama Beach
