---
title: "Making plots for linear regression ppt"
format: html
---



# Setup

Path

```{r}
setwd("week 3/Wk3-slides")
```

Libraries
```{r}

library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
library(patchwork)

library(easystats)
```

Datasets: Examples using iris and generated data

```{r}
# Regression exmaple
data_iris_reg <- iris |>
  filter(Species != "setosa") |>
  select(x = Sepal.Length, y = Sepal.Width, species = "Species")

# Anova exmaple
data_iris_anova_2grp <- iris |>
  filter(Species %in% c("setosa", "versicolor")) |>
  select(x = Species, y = Sepal.Width)

data_iris_anova_3grp <- iris |> select(x = Species, y = Sepal.Width)

## Generate data for simple example on intro slides
n <- 20
B0 <- 2
B1 <- 1.5

generate_data_reg <- function(n, B0, B1, sigma, lab, seed) {
  set.seed(seed)
  tibble(
    x = runif(n, 0, 4),
    y_hat = B0 + B1 * x,
    y = y_hat + rnorm(n, sd = sigma),
    error = lab
  )
}

# plots generated with seed = 110
data_gen_1_5 <- generate_data_reg(n, B0, B1, 1.1, "med", 110) |>
  mutate(y_bar = mean(y))
# saveRDS(data, "output/data_gen_1_5.rds")

## Generate data for ANOVA
generate_data_anova <- function(n, B0, B1, sigma, lab, seed) {
  set.seed(seed)
  tibble(
    x = rep(c(0, 1), n / 2),
    y_hat = B0 + B1 * x,
    y = y_hat + rnorm(n, sd = sigma),
    error = lab
  ) |>
    mutate(x = letters[x + 1] |> as.factor())
}

data_gen_a1 <- generate_data_anova(n, B0, B1 = 4, 1.1, "med", 110)

```

## Intro slides visuals - plot data and fitted line

Regression 
```{r}
## Fit model
fit <- lm(y ~ x, data = data_gen_1_5)

# basic scatter
p0 <-
  ggplot(data_gen_1_5, aes(x = x, y = y)) +
  geom_point(size = 2, col = "red") +
  theme_classic() +
  ylim(c(0, 10))

# format for small plot
p1 <- p0 +
  theme(
    panel.background = element_rect(fill = "transparent", color = NA), # Panel background
    plot.background = element_rect(fill = "transparent", color = NA), # Plot background
    axis.title.x = element_text(margin = margin(t = 10)), # Adds space above X-axis title
    axis.title.y = element_text(margin = margin(r = 10)), # Adds space to the right of Y-axis title
    axis.text = element_blank(), # Axis tick labels
    axis.title = element_text(size = 18), # Axis titles
    plot.title = element_text(size = 18, face = "bold") # Plot title
  )

#ggsave("output/scatter.png", p1, height = 2, width = 2, bg = "transparent")

## fitted line no ribbon, large size
p2a <-
  p0 +
  geom_smooth(method = "lm", col = "#eb8240", se = FALSE) +
  theme(
    panel.background = element_rect(fill = "transparent", color = NA), # Panel background
    plot.background = element_rect(fill = "transparent", color = NA), # Plot background
    axis.title.x = element_text(margin = margin(t = 10)), # Adds space above X-axis title
    axis.title.y = element_text(margin = margin(r = 10)), # Adds space to the right of Y-axis title
    #        axis.text = element_blank(),    # Axis tick labels
    axis.title = element_text(size = 18), # Axis titles
    plot.title = element_blank() # Plot title
  )

## fitted line - with ribbon, large size
p2b <- p0 +
  geom_smooth(method = "lm", col = "#eb8240") +
  theme(
    panel.background = element_rect(fill = "transparent", color = NA), # Panel background
    plot.background = element_rect(fill = "transparent", color = NA), # Plot background
    axis.title.x = element_text(margin = margin(t = 10)), # Adds space above X-axis title
    axis.title.y = element_text(margin = margin(r = 10)), # Adds space to the right of Y-axis title
    #        axis.text = element_blank(),    # Axis tick labels
    axis.title = element_text(size = 18), # Axis titles
    plot.title = element_text(size = 18, face = "bold") # Plot title
  )

#ggsave("output/scatter_lm.png", p2a, height = 2.5, width = 3.5, bg = "transparent")
#ggsave("output/scatter_lm_ribbon.png", p2b, height = 2.5, width = 3.5, bg = "transparent")

```

## Intro slide - Regression and ANOVA

use regression plot from above 

Now make Anova plot 
```{r}
## Fit model
fit <- lm(y ~ x, data = data_gen_a1)
means <- estimate_means(fit, by = c("x")) |> mutate(x = as.factor(x))

# basic scatter
p0a <-
  ggplot(data_gen_a1, aes(x = x, y = y)) +
  geom_errorbar(data = means, aes(ymin = CI_low, ymax = CI_high, y = Mean), width = 0, color = "darkgray", size = 5, alpha = 0.8) + # Confidence intervals
  # geom_point(data = means, aes(y = Mean), size = 6, color = "#eb8240") + # Estimated means
  geom_segment(data = means, aes(x = as.numeric(x) - 0.2, xend = as.numeric(x) + 0.2, y = Mean, yend = Mean), color = "#eb8240", size = 2) + # Estimated means
  scale_x_discrete() + # Keep x-axis categorical
  geom_jitter(color = "red", width = 0.1) +
  theme_classic() + # Style plot
  ylim(c(0, 10)) +
  labs(title = "Categorical x")


# format for small plot
p1a <- p0a +
  theme(
    panel.background = element_rect(fill = "transparent", color = NA), # Panel background
    plot.background = element_rect(fill = "transparent", color = NA), # Plot background
    axis.title.x = element_text(margin = margin(t = 10)), # Adds space above X-axis title
    axis.title.y = element_text(margin = margin(r = 10)), # Adds space to the right of Y-axis title
    # axis.text = element_blank(), # Axis tick labels
    axis.title = element_text(size = 18), # Axis titles
    plot.title = element_text(size = 18, face = "bold") # Plot title
  )

p_joint <- p1a / (p2b + labs(title = "Continuous x"))

ggsave("output/scatter_anova_regression.png", p_joint, height = 5, width = 4, bg = "transparent")
```

## Actual output in slides generated with iris

Note this output is different to above

```{r}
fit <- lm(y ~ x, data = data_iris_reg)

summary(fit)

parameters(fit)

report(fit)

# Estimate means and precited range
means <- estimate_means(fit, by = "x")
px <- means |> plot()

ggsave("output/scatter_estimate_means.png", px, height = 2.5, width = 3.5, bg = "transparent")

# simple plot with raw data points
ggplot(means, aes(x = x, y = Mean)) +
  geom_point(data = data_iris_reg, aes(x = x, y = y), color = "red") + # Raw data
  geom_line(size = 3, color = "orange") + # Estimated means
  theme_classic() + # Style plot
  labs(
    title = "Estimated Means with Raw Data"
  )

# Detailed Plot with raw data points
ggplot(means, aes(x = x, y = Mean)) +
  geom_ribbon(data = predicted, aes(ymin = CI_low, ymax = CI_high, y = Predicted), fill = "grey", alpha = 0.5) + # Confidence intervals
  geom_ribbon(aes(ymin = CI_low, ymax = CI_high), width = 0, fill = "orange", size = 3, alpha = 0.5) + # Confidence intervals
  geom_point(data = data_iris_reg, aes(x = x, y = y), color = "red") + # Raw data
  geom_line(size = 3, color = "orange") + # Estimated means
  theme_classic() + # Style plot
  labs(
    title = "Estimated Means with Raw Data"
  )
```


## demo Maximising likelihood

 create a 2d plot show residual sum of squares for different slope and intercept values
```{r}
# Custom RSS function
calculate_rss <- function(intercept, slope, data) {
  y_pred <- intercept + slope * data$x
  residuals <- data$y - y_pred
  sum(residuals^2)
}

# Create a grid of slope and intercept values
data_rss_grid <-
  expand_grid(intercept = seq(-5, 10, length.out = 100), slope = seq(-5, 8, length.out = 100)) |>
  mutate(rss = map2_dbl(intercept, slope, ~ calculate_rss(.x, .y, data = data_gen_1_5)))

# Plot RSS as a tile plot
p_rss <-
  ggplot(data_rss_grid, aes(x = intercept, y = slope, z = log(rss), fill = log(rss))) +
  geom_tile() +
  labs(
    title = "Residual Sum of Squares (RSS)",
    x = "B0",
    y = "B1",
    fill = "RSS"
  ) +
  geom_point(data = tibble(intercept = B0, slope = B1, rss = NA), col = "red") +
  scale_fill_viridis_c(option = "magma") + # "magma", "inferno", "plasma", "viridis", "cividis", etc.
  theme_classic() +
  theme(
    axis.title.x = element_text(margin = margin(t = 10)), # Adds space above X-axis title
    axis.title.y = element_text(margin = margin(r = 10)), # Adds space to the right of Y-axis title
    axis.text = element_blank(), # Axis tick labels
    axis.title = element_text(size = 18), # Axis titles
    plot.title = element_text(size = 18, face = "bold") # Plot title
  )

ggsave("output/rss.png", p_rss, height = 2, width = 2, bg = "transparent")
```

Create a gif of poor fits

```{r}
# Create a grid of values

make_plot <- function(b0, b1) {
  data_gen_1_5 |>
    mutate(y_bar = pmin(b0 + b1 * x, 15)) |>
    ggplot(aes(x = x, y = y)) +
    geom_point(size = 2, col = "red") +
    geom_abline(intercept = b0, slope = b1, col = "#eb8240") +
    geom_segment(aes(x = x, xend = x, y = y, yend = y_bar), col = "purple") +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    ylim(c(0, 15)) +
    theme_classic() +
    # labs(title = sprintf("B0 = %0.1f, B1 = %0.1f", b0, b1)) +
    theme(
      #      panel.background = element_rect(fill = "transparent", color = NA), # Panel background
      #      plot.background = element_rect(fill = "transparent", color = NA), # Plot background
      axis.title.x = element_text(margin = margin(t = 10)), # Adds space above X-axis title
      axis.title.y = element_text(margin = margin(r = 10)), # Adds space to the right of Y-axis title
      axis.text = element_blank(), # Axis tick labels
      axis.title = element_text(size = 18), # Axis titles
      plot.title = element_text(size = 18, face = "bold")
    )
}

make_plot(5, 4)

data_path <-
  tibble(
    intercept = seq(5, B0, length.out = 100),
    slope = seq(5, B1, length.out = 100)
  ) |>
  mutate(plot = map2(intercept, slope, ~ make_plot(.x, .y)))

# Save plots as png
walk2(data_path$plot, seq_along(data_path$plot), ~ ggsave(sprintf("output/MLE_path/resid_plot_%03d.png", .y), .x, height = 3, width = 5))

# make gif from pngs
system("convert -delay 10 -loop 10 output/MLE_path/resid_plot_*.png output/MLE_path/resid_movie.gif")

# make it smaller
system("convert resid_movie.gif -resize 50% -coalesce -set delay 3 -layers Optimize resid_movie_opt.gif")

```


## Plot lines with different slope

```{r}

z <- c(c(-1, 0, 0.5, 2.5))
labs <- c("negative", "flat", "positive", "stronger positive")

data2 <- map2_df(z, factor(labs, levels = labs), ~generate_data_reg(n, B0, .x, 1.1, .y, 110))

pz <- 
    ggplot(data2, aes(x = x, y = y)) +
    geom_point(size=2, col= "red") +
    theme_classic() +
    ylim(c(-5, 12)) +
    geom_smooth(method= "lm", col ="#eb8240", se = FALSE) +
    theme(
        panel.background = element_rect(fill = "transparent", color = NA), # Panel background
        plot.background = element_rect(fill = "transparent", color = NA),  # Plot background
        axis.title.x = element_text(margin = margin(t = 10)),  # Adds space above X-axis title
        axis.title.y = element_text(margin = margin(r = 10)),   # Adds space to the right of Y-axis title
#        axis.text = element_blank(),    # Axis tick labels
        axis.title = element_text(size = 18),   # Axis titles
        plot.title = element_blank(), # Plot title
        strip.background = element_blank(),  # Removes the box
        strip.text = element_text(face = "bold"), # Optional: Style the text
axis.text.x = element_text()
    ) + facet_wrap(~error, ncol=1) 

pz

ggsave("output/scatter_lm_panel_slopes.png", pz, height = 5, width = 3, bg = "transparent")

```

## Confidence intervals

Variation among samples

```{r}


data2 <- map_df(1:4, ~generate_data_reg(5, B0, B1, 1.1, .x, .x + 20))

pz <- 
    ggplot(data2, aes(x = x, y = y)) +
    geom_abline(intercept = B0, slope = B1, col = "black") +
    geom_point(size=2, col= "red") +
    theme_classic() +
    ylim(c(0, 10)) +
    xlim(c(0, 4.5)) +
    geom_smooth(method= "lm", col ="#eb8240", se = FALSE) +
    theme(
        panel.background = element_rect(fill = "transparent", color = NA), # Panel background
        plot.background = element_rect(fill = "transparent", color = NA),  # Plot background
        axis.title.x = element_text(margin = margin(t = 10)),  # Adds space above X-axis title
        axis.title.y = element_text(margin = margin(r = 10)),   # Adds space to the right of Y-axis title
#        axis.text = element_blank(),    # Axis tick labels
        axis.title = element_text(size = 18),   # Axis titles
        plot.title = element_blank(), # Plot title
        strip.background = element_blank(),  # Removes the box
        strip.text = element_blank(), # Optional: Style the text
axis.text.x = element_text()
    ) + facet_wrap(~error, ncol=1) 

pz

ggsave("output/scatter_lm_panel_samples.png", pz, height = 5, width = 3, bg = "transparent")

```

Variation with sample size

```{r}
nz <- c(5,10, 20, 40)
labs <- sprintf("n = %s", nz)
labs <- factor(labs, levels = labs)

data2 <- map2_df(nz, labs, ~generate_data_reg(.x, B0, B1, 1.1, .y, 8)) 

pz <- 
    ggplot(data2, aes(x = x, y = y)) +
    geom_abline(intercept = B0, slope = B1, col = "black") +
    geom_point(size=2, col= "red") +
    theme_classic() +
    ylim(c(0, 10)) +
    xlim(c(0, 4.5)) +
    geom_smooth(method= "lm", col ="#eb8240", se = TRUE) +
    theme(
        panel.background = element_rect(fill = "transparent", color = NA), # Panel background
        plot.background = element_rect(fill = "transparent", color = NA),  # Plot background
        axis.title.x = element_text(margin = margin(t = 10)),  # Adds space above X-axis title
        axis.title.y = element_text(margin = margin(r = 10)),   # Adds space to the right of Y-axis title
#        axis.text = element_blank(),    # Axis tick labels
        axis.title = element_text(size = 18),   # Axis titles
        plot.title = element_blank(), # Plot title
        strip.background = element_blank(),  # Removes the box
        strip.text = element_text(face = "bold"), # Optional: Style the text
axis.text.x = element_text()
    ) + facet_wrap(~error, ncol=1) 

pz

ggsave("output/scatter_lm_panel_n.png", pz, height = 5, width = 3, bg = "transparent")

```

## Evaluating fit - variance explained

```{r}
## Fit model
fit <- lm(y ~ x, data = data_gen_1_5)

means <- estimate_means(fit, by = "x", data = data_gen_1_5)

data_plot <- data_gen_1_5 |>
  mutate(
    y_fit = predict(fit),
    y_bar = mean(y)
  )

# basic scatter

p_tot <-
  ggplot(data_plot, aes(x = x, y = y)) +
  theme_classic() +
  ylim(c(0, 10)) +
  geom_line(aes(y = y_bar), linetype = "dashed", col = "#eb8240") +
  geom_segment(aes(x = x, xend = x, y = y, yend = y_bar), col = "purple") +
  geom_point(size = 2, col = "red") +
  labs(title = "Total variation", x = "") +
  theme(
    panel.background = element_rect(fill = "transparent", color = NA), # Panel background
    plot.background = element_rect(fill = "transparent", color = NA), # Plot background
    #axis.title.x = element_text(margin = margin(t = 10)), # Adds space above X-axis title
    axis.title.y = element_text(margin = margin(r = 10)), # Adds space to the right of Y-axis title
    axis.text = element_blank(), # Axis tick labels
    axis.title = element_text(size = 18), # Axis titles
    plot.title = element_text(size = 18, face = "bold") # Plot title
  )

p_model <-
  ggplot(data_plot, aes(x = x, y = y)) +
  theme_classic() +
  ylim(c(0, 10)) +
  geom_line(aes(y = y_bar), linetype = "dashed", col = "#eb8240") +
  geom_line(aes(y = y_fit), col = "#eb8240") +
  geom_segment(aes(x = x, xend = x, y = y_fit, yend = y_bar), col = "purple") +
  geom_point(size = 2, col = "red") +
  #geom_point(size = 2, col = "#eb8240", aes(y = y_fit)) +
  labs(title = "Modelled variation", x = "") +
  theme(
    panel.background = element_rect(fill = "transparent", color = NA), # Panel background
    plot.background = element_rect(fill = "transparent", color = NA), # Plot background
    #axis.title.x = element_text(margin = margin(t = 10)), # Adds space above X-axis title
    axis.title.y = element_text(margin = margin(r = 10)), # Adds space to the right of Y-axis title
    axis.text = element_blank(), # Axis tick labels
    axis.title = element_text(size = 18), # Axis titles
    plot.title = element_text(size = 18, face = "bold") # Plot title
  )

p_resid <-
  ggplot(data_plot, aes(x = x, y = y)) +
  theme_classic() +
  ylim(c(0, 10)) +
  geom_line(aes(y = y_fit), col = "#eb8240") +
  geom_segment(aes(x = x, xend = x, y = y, yend = y_fit), col = "purple") +
  geom_point(size = 2, col = "red") +
  labs(title = "Residual variation") +
  theme(
    panel.background = element_rect(fill = "transparent", color = NA), # Panel background
    plot.background = element_rect(fill = "transparent", color = NA), # Plot background
    axis.title.x = element_text(margin = margin(t = 10)), # Adds space above X-axis title
    axis.title.y = element_text(margin = margin(r = 10)), # Adds space to the right of Y-axis title
    axis.text = element_blank(), # Axis tick labels
    axis.title = element_text(size = 18), # Axis titles
    plot.title = element_text(size = 18, face = "bold") # Plot title
  )

p_all <- p_tot / p_model / p_resid

ggsave("output/variance_partition.png", p_all, height = 6, width = 3.5, bg = "transparent")
```

## Plot lines with R2 values

```{r}
z <- c(0, 0.25, 2, 5)
labs <- z |> as.character()

data2 <- map2_df(z, factor(labs, levels = labs), ~ generate_data_reg(n, B0, .x, 1.1, .y, 110))

data2 |>
  nest(data = c(x, y, y_hat)) |>
  mutate(
    fit = map(data, ~lm(y ~ x, data = .x)),
    summary = map(fit, ~summary(.x)),
    f = map_dbl(summary, ~.x$fstatistic[1]),
    r2 = map_dbl(summary, ~.x$r.squared),
    p = map_dbl(summary, ~ pf(.x$fstatistic[1], .x$fstatistic[2], .x$fstatistic[3], lower.tail = FALSE))
    )

pz <-
  ggplot(data2, aes(x = x, y = y)) +
  geom_point(size = 2, col = "red") +
  theme_classic() +
  # ylim(c(-5, 25)) +
  geom_smooth(method = "lm", col = "#eb8240", se = FALSE) +
  theme(
    panel.background = element_rect(fill = "transparent", color = NA), # Panel background
    plot.background = element_rect(fill = "transparent", color = NA), # Plot background
    axis.title.x = element_text(margin = margin(t = 10)), # Adds space above X-axis title
    axis.title.y = element_text(margin = margin(r = 10)), # Adds space to the right of Y-axis title
    #        axis.text = element_blank(),    # Axis tick labels
    axis.title = element_text(size = 18), # Axis titles
    plot.title = element_blank(), # Plot title
    strip.background = element_blank(), # Removes the box
    strip.text = element_text(face = "bold"), # Optional: Style the text
    axis.text.x = element_text()
  ) +
  facet_wrap(~error, ncol = 1, scales = "free_y")

pz

ggsave("output/scatter_lm_panel_r2_f_p.png", pz, height = 5, width = 3, bg = "transparent")

```

## Evaluating fit - F distribution

```{r}
df1 <- 1
df2 <- 18
F_stat <- summary(fit)$fstatistic[1]
p_val <- pf(F_stat, df1, df2, lower.tail = FALSE)
F_crit <- qf(p = 0.95, df1 = df1, df2 = df2)

x_vals <- seq(0.1, 100, length.out = 500)
y_vals <- df(x_vals, df1, df2)
p_vals <- pf(x_vals, df1, df2, lower.tail = FALSE)
df_data <- tibble(x = x_vals, y = y_vals, p = p_vals)

p <- ggplot(df_data, aes(x, y)) +
  geom_line(color = "darkgrey", size = 1) +
  labs(
    title = paste("F (df1 =", df1, ", df2 =", df2, ")"),
    x = "F",
    y = "density"
  ) +
    geom_point(data  = tibble(x = F_stat, y = pf(F_stat, df1, df2, lower.tail = FALSE), col="red")) +
  theme_classic() +
  scale_x_log10() +
  theme(
      panel.background = element_rect(fill = "transparent", color = NA), # Panel background
      plot.background = element_rect(fill = "transparent", color = NA),  # Plot background
      axis.title.x = element_text(margin = margin(t = 8), size = 18),  # Adds space above X-axis title
      axis.title.y = element_text(margin = margin(r = 10),size = 12),   # Adds space to the right of Y-axis title
      axis.text.x = element_text(size = 10),    # Axis tick labels
      axis.text.y = element_blank(),    # Axis tick labels
      plot.title = element_blank() # Plot title
  )

ggsave("output/F-dist.png", p, height = 2, width = 2, bg = "transparent")

```


## pval


Variation among samples when B0 = 0

```{r}


data2 <- map_df(1:4, ~generate_data_reg(5, B0, 0, 1.1, .x, .x + 20))

pz <- 
    ggplot(data2, aes(x = x, y = y)) +
    geom_abline(intercept = B0, slope = 0, col = "black") +
    geom_point(size=2, col= "red") +
    theme_classic() +
    ylim(c(0, 10)) +
    xlim(c(0, 4.5)) +
    geom_smooth(method= "lm", col ="#eb8240", se = FALSE) +
    theme(
        panel.background = element_rect(fill = "transparent", color = NA), # Panel background
        plot.background = element_rect(fill = "transparent", color = NA),  # Plot background
        axis.title.x = element_text(margin = margin(t = 10)),  # Adds space above X-axis title
        axis.title.y = element_text(margin = margin(r = 10)),   # Adds space to the right of Y-axis title
#        axis.text = element_blank(),    # Axis tick labels
        axis.title = element_text(size = 18),   # Axis titles
        plot.title = element_blank(), # Plot title
        strip.background = element_blank(),  # Removes the box
        strip.text = element_blank(), # Optional: Style the text
axis.text.x = element_text()
    ) + facet_wrap(~error, ncol=1) 

pz

ggsave("output/scatter_lm_panel_samples_NULL.png", pz, height = 5, width = 3, bg = "transparent")

```


# ANOVA 

```{r}
fit <- lm(y ~ x, data = data_iris_anova_3grp)
means <- estimate_means(fit, by = "x")

summary(fit)

anova(fit)
report(fit)

estimate_contrasts(fit, contrast = "x")


# simple plot
means |> plot(line = list(size = 0), col = list(col = "#eb8240")) +
  geom_jitter(data = data_iris_anova_3grp, aes(x = x, y = y), color = "red", width = 0.05) + # Confidence intervals
  theme_classic()

# Plot with raw data points
pa1 <-
  ggplot(means, aes(x = x, y = Mean)) +
  geom_jitter(data = data_iris_anova_3grp, aes(x = x, y = y), color = "red", width = 0.1) + # Raw data
  geom_errorbar(aes(ymin = CI_low, ymax = CI_high), width = 0, color = "grey", size = 3) + # Confidence intervals
  geom_point(size = 5, color = "orange") + # Estimated means
  theme_classic() + # Style plot
  labs(y = "y") +
  theme(
    panel.background = element_rect(fill = "transparent", color = NA), # Panel background
    plot.background = element_rect(fill = "transparent", color = NA), # Plot background
    axis.title.x = element_text(margin = margin(t = 10)), # Adds space above X-axis title
    axis.title.y = element_text(margin = margin(r = 10)), # Adds space to the right of Y-axis title
    #   axis.text = element_blank(), # Axis tick labels
    axis.title = element_text(size = 18), # Axis titles
    plot.title = element_text(size = 18, face = "bold") # Plot title
  )

ggsave("output/scatter_anova.png", pa1, height = 3.5, width = 5, bg = "transparent")




```




## GLM vs lm

```{r}
# Gaussian glm (equivalent to lm)

lm_fit <- lm(mpg ~ hp + wt, data = mtcars)
glm_fit <- glm(mpg ~ hp + wt, data = mtcars, family = gaussian)
TMB_fit <- glmmTMB::glmmTMB(mpg ~ hp + wt, data = mtcars)

# summary
summary(lm_fit)
summary(glm_fit)  


# Traditional ANOVA table
anova(lm_fit)
anova(glm_fit)  

parameters(lm_fit)
parameters(glm_fit)
parameters(TMB_fit)

report(lm_fit)
report(glm_fit)


# categorical only
lm_fit2 <- lm(mpg ~ as.factor(gear), data = mtcars)
glm_fit2 <- glm(mpg ~as.factor(gear), data = mtcars, family = gaussian)

anova(lm_fit2)
anova(glm_fit2)  

parameters(lm_fit2)
parameters(glm_fit2)

# mix

lm_fit3 <- lm(mpg ~ hp*as.factor(gear), data = mtcars)
glm_fit3 <- glm(mpg ~  hp*as.factor(gear), data = mtcars)
TMB_fit2 <- glmmTMB::glmmTMB(mpg ~  as.factor(gear), data = mtcars)
TMB_fit3 <- glmmTMB::glmmTMB(mpg ~  hp*as.factor(gear), data = mtcars)

summary(lm_fit3)
summary(glm_fit3)
summary(TMB_fit3)

anova(lm_fit3)
anova(glm_fit3)  
anova(TMB_fit2, TMB_fit3)  

parameters(lm_fit3)
parameters(glm_fit3)
parameters(TMB_fit3)

performance(lm_fit3)
performance(glm_fit3)
performance(TMB_fit3)

report(lm_fit3)
report(glm_fit3)
report(TMB_fit3)  # some minor issues here

check_model(lm_fit3)
check_model(glm_fit3)
check_model(TMB_fit3)

estimate_contrasts(lm_fit3, contrast = "gear")
estimate_contrasts(glm_fit3, contrast = "gear")
estimate_contrasts(TMB_fit3, contrast = "gear")

## two way anova
glm_fit4 <- lm(mpg ~ as.factor(hp)*as.factor(gear), data = mtcars)
TMB_fit4 <- glmmTMB::glmmTMB(mpg ~ as.factor(hp)*as.factor(gear), data = mtcars)

```


# Examples

## Regression - Towers 2024

```{r}
# data_towers <- readRDS("data/ignore/Towers_2024_woody_traits_climate.rds") |>
#   filter(growth_form == "Woody") |>
#   select(cell,
#     temperature = Temp, rainfall = Prec,
#     # leaf_area, wood_density,
#     leaf_mass_per_area
#   ) |>
#   mutate(
#     log10_rainfall = log10(rainfall),
#     log10_leaf_mass_per_area = log10(leaf_mass_per_area)
#   ) |>
#   drop_na()
# data_towers |> saveRDS("data/Towers2024.rds")

data_towers <- readRDS("data/Towers2024.rds")

fit <- lm(log10_leaf_mass_per_area ~ log10_rainfall, data = data_towers)

means <- estimate_means(fit, by = "log10_rainfall")

summary(fit)

p_results <-
  ggplot(data_towers, aes(x = log10_rainfall, y = log10_leaf_mass_per_area)) +
  geom_point(size = 2, col = "red", alpha = 0.2, stroke = 0) +
  theme_classic() +
  labs(
    title = "Response of woody-plants to rainfall",
    x = "log10 Rainfall",
    y = "log10 Leaf Mass per Area"
  ) +
  theme(
    panel.background = element_rect(fill = "transparent", color = NA), # Panel background
    plot.background = element_rect(fill = "transparent", color = NA), # Plot background
    axis.title.x = element_text(margin = margin(t = 10)), # Adds space above X-axis title
    axis.title.y = element_text(margin = margin(r = 10)), # Adds space to the right of Y-axis title
    axis.title = element_text(size = 14), # Axis titles
    plot.title = element_text(size = 14, face = "bold") # Plot title
  )

p_results_lm <-
  p_results +
  geom_ribbon(data = means, aes(ymin = CI_low, ymax = CI_high, y = Mean), fill = "grey") +
  geom_line(data = means, aes(y = Mean), col = "#eb8240", size = 1)

ggsave("output/example_scatter.png", p_results, height = 4, width = 5, bg = "transparent")


ggsave("output/example_lm.png", p_results, height = 4, width = 5, bg = "transparent")

```
