---
title: "Introduction to data analysis"
output: html_notebook
---

Load packages required

```{r}
library(ggplot2)
library(dplyr)
```

If you get an error message, then you may need to install the packages onto your machine. You can do this with


Q1 Samples and populations

Import the data file (this code will work if the data file is saved in the same folder or directory as this notebook file - the one titled Introduction_sample.Rmd)

```{r} 
Population <- read.csv(file = "Population.csv", header = T)
```

Create a frequency histogram of the size variable

```{r}
ggplot(Population, aes(size)) + geom_histogram()
```

Q1a) Calculate the mean of the size variable


```{r}
mean(Population$Size)
```

Q1b) Calculate the variance of your population


```{r}
pop.var <- mean((Population$size-mean(Population$size))^2)
pop.var
```


Q1c) Calculate the standard deviation of your population

```{r}
sqrt(pop.var)
```

Sample the population (5, 10 20 and 50 measures taken randomly)

```{r}
Sample5 <- sample_n(Population, 5)
Sample10 <- sample_n(Population, 10)
Sample20 <- sample_n(Population, 20)
Sample50 <- sample_n(Population, 50)
```


Q1d) Calculate the mean, variance and standard deviation for each of these samples. 

```{r}
mean(Sample5$size)
sd(Sample5$size)
var(Sample5$size)

mean(Sample10$size)
sd(Sample10$size)
var(Sample10$size)

mean(Sample20$size)
sd(Sample20$size)
var(Sample20$size)

mean(Sample50$size)
sd(Sample50$size)
var(Sample50$size)
```


```{r}
chisq.test(x = c(153,20,138), p = c(100/311,89/311,122/311))
```

```{r}
Q18 <- c(874, 745, 1002, 554, 837, 901, 845, 912, 698, 712)
t.test(Q18, alternative = "two.sided", mu = 670)
```

```{r}
Q15 <-c (225, 341, 240, 234, 270, 334, 330, 248, 169, 336)
t.test(Q15, alternative = "greater", mu = 300)
```
```{r}
Crab <- read.csv("Fiddler_crab.csv", header = T)
t.test(Duration ~ Side, data = Crab, paired = TRUE, alternative = "less")
t.test(Duration ~ Side, data = Crab, paired = TRUE, alternative = "greater")
t.test(Duration ~ Side, data = Crab, paired = TRUE, alternative = "two.sided")
```
```{r}
chisq.test(x = c(8,14,2,7,24,1), p = c(0.05, 0.25, 0.24, 0.31, 0.1, 0.05))
```
```{r}
contam <- c(225, 341, 240, 234, 270, 334, 330, 248, 169, 336)
mean(contam) - qt(0.975, df = 9) * sd(contam)/sqrt(10)
```
Q45
B) Conduct the test, and give the value of the test statistic {1:NUMERICAL:=0.23874:0.01} and P value

{1:NUMERICAL:=0.8208:0.001}


```{r}
cows <- data.frame(Cow = c(c(1:6), c(1:6)), source = c(rep("normal", 6), rep("elevated", 6)),
copper = c(0.59, 0.63, 0.25, 0.84, 0.48, 0.63, 0.64, 0.54, 0.35, 0.42, 0.71, 0.63))
t.test(copper~ source, data = cows, paired = TRUE, alternative="two.sided")
```

Q18 Linear regression

```{r}
students <- data.frame(Name = c("Jack", "Emma","Aiden","Sharon","Gabriela"), Hours = c(6,8.6, 8,7,4.2),Grade = c(53,84,92,71,42))
```

```{r}
mod18 <- lm(Grade ~ Hours,data = students)
summary(mod18)
```

```{r}
Mink <- read.csv("Mink.csv", header = T)
Mink.nested <- aov(Voles~ Treatment + Area + Area%in%Treatment, data = Mink)
plot(Mink.nested)
summary(Mink.nested)
```

```{r}
T <-(1416.1-178.9)/20
A <- (178.9-36.9)/10
E <- 36.9

tp <- T/(T+A+E)*100
Ap <- A/(T+A+E)*100
Ep <- E/(T+A+E)*100
```



