---
title: "Week 1-1 Data manipulation and visualisation"
output: html_document()
#   moodlequiz::moodlequiz:
#     replicates: 1
# moodlequiz:
#   category: "Week 1-2 data wrangle + viz"
# editor_options:
#   chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# remotes::install_github("numbats/moodlequiz")
library(moodlequiz)
library(here)

# Libraries for prac
library(tidyverse)
library(janitor)

```

## Overview

Professor Angela Moles, a professor here at UNSW, used an open-source dataset of maximum height data for 7084 plant species and site combinations to provide the first global, cross-species quantification of the latitudinal gradient in plant height. (You can read more about the original study at [doi:10.1111/j.1365-2745.2009.01526.x](doi.org/10.1111/j.1365-2745.2009.01526.x) or in `docs/Journal of Ecology - 2009 - Moles - Global patterns in plant height.pdf`)

We are going to use this data to do a bit of preliminary analyses and see if we can answer two questions:

**1. How does plant height vary across a latitudinal gradient (from tropics to the poles)?**

**2. Are there taller plants at higher altitudes?**

**Activities:**

- Read in data and wrangle it suited for visualisation

- Create data visualisation that can address the questions

- Draw some initial conclusions from the plots you create

## Setting up

```{r}
# install.packages("tidyverse")
library(tidyverse)
```

## Assembling the Data

Read in the dataset `plantHeightSingleSpp.csv` from the folder `data/plant_height/`:

```{r}
# plant_height <- read_csv("week 1/Wk1-2-moodle/data/plant_height/plantHeightSingleSpp.csv")
plant_height <- read_csv("data/plant_height/plantHeightSingleSpp.csv")
```

WOAH there are a LOT of variables in this data set. Sometimes it's nice to select only the columns you need for your analysis. This can make things easier in the long-run. Luckily, `dplyr` can help with that!

**Question: Which columns are relevant to these 2 questions? Select all that apply**

<!-- TODO `r cloze(c("boxplot", "histogram", "violin"), possible_choices)` -->

Now, create a new data frame with only the relevant columns using `select()`

```{r}
plant_height_subset <- plant_height |> select(Genus_species, height, alt, lat)
```

## Check the distribution for the data

Check the distribution of your data to make sure there are no potential outliers in your key variables. What plot would be good for this check? 

<!-- TODO: `r cloze(c("boxplot", "histogram", "violin"), possible_choices)` -->

```{r}
plant_height_subset |> 
  ggplot(aes(x = height)) + 
  geom_histogram()
```

<!-- TODO Is this data left skewed or right skewed? -->

<!-- TODO Is the extreme value a true outlier? Which species is this? Google search and see if you can verify this yourself -->

```{r}
plant_height_subset |> 
  filter(height > 60)
```

## Data transformation

It is common to transform your data when its heavily skewed as in `height`, can you `log10()` transform this variable and assign this update into a new data frame.

```{r}
plant_height_transformed <- plant_height_subset |> 
  mutate(log10_height = log10(height))
```

## Visualisation

<!-- TODO: Question: What should be the x variable and what should be the y variable? -->

We'll first create a `ggplot` showing `lat` on the `x`-axis and `height` on the `y`-axis. Then, we'll create the same figure but using the `log10` transformed data to see how it changes the patterns.

> Create a plot to visualise the relationship between latitude and height.

<!-- TODO **Question: What type of plot is suited to visualise these two variables?** -->

```{r}
ggplot(plant_height_transformed, aes(x = lat, y =height)) + 
  geom_point()
```

Let's take a look at the line of best fit! To add this to our figure, use `geom_smooth(method="lm")`.

```{r}
ggplot(plant_height_transformed, aes(x = lat, y = height)) + 
  geom_point() +
  geom_smooth(method="lm")
```

Now create the same plot but using your `log10()` transformed variable instead of `height` and include the line of best fit.

```{r}
ggplot(plant_height_transformed, aes(x = lat, y = log10_height)) + 
  geom_point() +
  geom_smooth(method="lm")
```

<!-- TODO **Question: 1. How does plant height vary across a latitudinal gradient (from tropics to the poles)?** -->

Plant height increases/decrease from tropics to poles

> Now, we are going to try to answer our second question; Does plant height increase as **altitude** increases?

## Altitude

We will continue to use the `log10` transformed `height`

Make a plot that can help us answer question 2. - add the line of best fit.

```{r}
ggplot(plant_height_transformed, aes(x = alt, y = log10_height)) + 
  geom_point() +
  geom_smooth(method="lm")
```

<!-- TODO **Question: 2. Are there taller plants at higher altitudes?** -->

There are taller/shorter trees in higher altitudes

## Prac reflection {type=essay}

<h3>Prac reflection</h3>

<!-- TODO  **Question: Which parts of the coding and the tidyverse you've encountered so far did you find:** -->

- most exciting?
- most challenging?
